---
title: "GLMM example 5"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(brms)
library(DHARMa)
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
owls = read_csv('../data/owls.csv', trim_ws=TRUE)
glimpse(owls)
owls <- mutate(owls,Nest=factor(Nest),FoodTreatment=factor(FoodTreatment),SexParent=factor(SexParent),NCalls=SiblingNegotiation) # to rename instead of add on extra variable use 'rename' function
# number of negotations that owl chicks make - calls - so response is count data (Poisson or negBinomial)

# TOP LEVEL: Nests - random level effect
# what things vary within the nest, what things vary between the nest
# NEXT LEVEL: FoodTreatment - each nest had both treatments
# Arrival Time - just recorded the time - continuous variable - measured multiple times per nest 
# BroodSize is covariate - to standardise number of calls for brood size - include it as an offset so no parameters are estimated for it - assume 1:1 relationship
```


# Exploratory data analysis

Model formula:
$$
y_i \sim{} \mathcal{Pois}(\lambda_i)\\
ln(\lambda_i) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of food treatment, sex of parent, arrival time (and various interactions) on the number of sibling negotiations.
Brood size was also incorporated as an offset.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual nests.

```{r exploratory plots}
# NCalls as a function of ArrivalTime coloured by FoodTreatment with a smoother added in to just visualise any trend
ggplot(owls,aes(y=NCalls,x=ArrivalTime,color=FoodTreatment))+
  geom_point()+geom_smooth(method='lm')+theme_bw()

# NCalls as a function of ArrivalTime coloured by FoodTreatment & separate graph for each nest 
ggplot(owls,aes(y=NCalls,x=ArrivalTime,color=FoodTreatment))+
  geom_point()+geom_smooth(method='lm')+
  facet_wrap(~Nest)+theme_bw()

ggplot(owls,aes(y=NegPerChick,x=ArrivalTime,color=FoodTreatment))+
  geom_point()+geom_smooth(method='lm')+
  facet_wrap(~Nest,scale='free_y')+
  theme_bw()

# ok so might want to fit a random slope & random slope random intercept model

# NCalls~FoodTreatment+scale(ArrivalTime)+offset(log(BroodSize))+(1|Nest)
# NCalls~FoodTreatment+scale(ArrivalTime)+offset(log(BroodSize))+(FoodTreatment|Nest)??
```

# FIT model owls.brmsP
```{r fit poisson model, cache=TRUE}
owls.form <- bf(NCalls~FoodTreatment+scale(ArrivalTime)+offset(log(BroodSize))+(1|Nest),family=poisson(link='log'))
# NB: scaling continuous variables speed up analyses
owls.brmsP <- brm(owls.form,data=owls,iter=5000,warmup=2000,chains=3,refresh=0,thin=5)
```

# VALIDATE model owls.brmsP
```{r chain characteristics}
mcmc_plot(owls.brmsP,type='trace')      #
mcmc_plot(owls.brmsP,type='acf_bar')    # autocorrelation
mcmc_plot(owls.brmsP,type='neff_hist')  # number of effectives
mcmc_plot(owls.brmsP,type='rhat_hist')  # chain convergence
```

```{r test for overdispersion of brmsP}
preds <- posterior_predict(owls.brmsP,nsamples=250,summary=FALSE)
owls.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = owls$NCalls,
                             fittedPredictedResponse = apply(preds,2,median))
plot(owls.resids) # BAHH this is NOT good - we have overdispersion (not uncommon for Poisson distributions) & outliers 
# could be overdispersed because model is too simplistic
# could be overdispersed because of zero inflation
```


```{r test for zeroinflation}
#  test for zero inflation
testZeroInflation(owls.resids) # yep there's a problem here with zero inflation => can account for this with Negative Binomial BUT this dataset has A LOT of zeros so we suspect Negative Binomial won't be able to handle this - instead we try a zero Inflation model 
```

# RE-FIT model owls.brmsZIP
```{r fit zero inflated poisson model, cache=TRUE}
owls.form1 <- bf(NCalls~FoodTreatment+scale(ArrivalTime)+offset(log(BroodSize))+(1|Nest), # count process - counts that it                                                                                              thinks are true
                 zi~1,                                                                    # detection process (can I detect                                                                                             or not)
                 family=zero_inflated_poisson(link='log'))
# zi~ 1 - not modelling any reasoning into my ability to detect - you could model zero inflation based on something like ArrivalTime or something else e.g. zi~scale(ArrivalTime) - this is unlikely to help the model run better in this case 
owls.brmsZIP <- brm(owls.form1,data=owls,iter=5000,warmup=2000,chains=3,refresh=0,thin=10)
```

# VALIDATE model owls.brmsZIP
```{r check chain characteristics brmsZIP}
mcmc_plot(owls.brmsZIP,type='trace')      #
mcmc_plot(owls.brmsZIP,type='acf_bar')    # autocorrelation
mcmc_plot(owls.brmsZIP,type='neff_hist')  # number of effectives
mcmc_plot(owls.brmsZIP,type='rhat_hist')  # chain convergence
```

```{r test for overdispersion brmsZIP}
preds <- posterior_predict(owls.brmsZIP,nsamples=250,summary=FALSE)
owls.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = owls$NCalls,
                             fittedPredictedResponse = apply(preds,2,median))
plot(owls.resids) # this is still not great..... so overdispersion probably needs to be dealt with negative binomial as well as zero inflation - so let's try fit a zero inflated negative binomial
```
# RE-FIT model owls.brmsZINB
```{r fit zero inflated negative binomial model, cache=TRUE}
owls.form2 <- bf(NCalls~FoodTreatment+scale(ArrivalTime)+offset(log(BroodSize))+(1|Nest), 
                 zi~scale(ArrivalTime),                                                                    
                 family=zero_inflated_negbinomial(link='log'))
owls.brmsZINB <- brm(owls.form2,data=owls,iter=5000,warmup=2000,chains=3,refresh=0,thin=10)
```

# VALIDATE model owls.brmsZINB
```{r check chain characteristics brmsZINB}
mcmc_plot(owls.brmsZINB,type='trace')      #
mcmc_plot(owls.brmsZINB,type='acf_bar')    # autocorrelation
mcmc_plot(owls.brmsZINB,type='neff_hist')  # number of effectives
mcmc_plot(owls.brmsZINB,type='rhat_hist')  # chain convergence
```

```{r test for overdispersion brmsZINB}
preds <- posterior_predict(owls.brmsZINB,nsamples=250,summary=FALSE)
owls.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = owls$NCalls,
                             fittedPredictedResponse = apply(preds,2,median))
plot(owls.resids) # this is STILL bad - so let's try adding in a random slope to increase the complexity of the model - maybe that will help 
```

# RE-FIT model owls.brmsZINB1
```{r fit zero inflated negative binomial model random intercept random slope, cache=TRUE}
# let's change the model one more time adding in random slope to see if that helps (i.e. the model might be too simple)
owls.form3 <- bf(NCalls~FoodTreatment+scale(ArrivalTime)+offset(log(BroodSize))+(scale(ArrivalTime)|Nest),  
                 zi~scale(ArrivalTime),                                                                    
                 family=zero_inflated_negbinomial(link='log'))
# you could do ((FoodTreatment+scale(ArrivalTime))|Nest) if you thought that was pattern with foodtreatment also, but it take a lot longer - depends on your exploratory plot what sort of patterns you were seeing
owls.brmsZINB1 <- brm(owls.form3,data=owls,iter=5000,warmup=2000,chains=3,refresh=0,thin=10)
```

# VALIDATE model owls.brmsZINB1
```{r test for overdispersion brmsZINB1}
preds <- posterior_predict(owls.brmsZINB1,nsamples=250,summary=FALSE)
owls.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = owls$NCalls,
                             fittedPredictedResponse = apply(preds,2,median))
plot(owls.resids) # this is STILL bad - so let's try adding in a random slope to increase the complexity of the model - maybe that will help 
```


# RE-FIT model owls.brmsZINB2
```{r fit zero inflated negative binomial model random intercept random slope brmsZINB2, cache=TRUE}
# let's change the model one more time by adding in an interaction to see if that helps
owls.form4 <- bf(NCalls~FoodTreatment*scale(ArrivalTime)+offset(log(BroodSize))+(scale(ArrivalTime)|Nest),  
                 zi~scale(ArrivalTime),                                                                    
                 family=zero_inflated_negbinomial(link='log'))
owls.brmsZINB2 <- brm(owls.form4,data=owls,iter=5000,warmup=2000,chains=3,refresh=0,thin=10)
```

# VALIDATE model owls.brmsZINB2
```{r test for overdispersion brmsZINB2}
preds <- posterior_predict(owls.brmsZINB2,nsamples=250,summary=FALSE)
owls.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = owls$NCalls,
                             fittedPredictedResponse = apply(preds,2,median))
plot(owls.resids) # this will have to do - at some point you have to be ok with perhaps some small issues as it probably won't make much difference to your conclusions 
```

# INVESTIGATE MODEL / TEST HYPOTHESES
```{r model summary}
# ok so we've chosen owls.brmsZINB2 as the model to proceed with - let's have a look at it a bit further:
tidyMCMC(owls.brmsZINB2$fit,conf.int=TRUE,conf.method='HPDinterval',rhat=TRUE,ess=TRUE)

# EXAMPLE INTERPRETATIONS OF MODEL SUMMARY RESULTS:
# b_zi_Intercept average NCalls (on log scale) for the deprived group at the average Arrival Time (& obviously averaged over the nests)
# b_FoodTreatmentSatiated	-0.274280665	: satiated chicks call less (conf interval indicates this difference is significant - so food treatment seems to have an effect)

# b_zi_scaleArrivalTime	0.336646073 - the chance of false zeros increases as arrival time increases - and this seems to be significant - more & more zeros as arrival time increased - & more and more of them you would consider as false zeros 
```

```{r plot model by making predictions first}
ggpredict(owls.brmsZINB2) %>% plot

owls.grid <- with(owls,list(FoodTreatment=levels(FoodTreatment),ArrivalTime=seq(min(ArrivalTime),max(ArrivalTime),len=100)))
owls.new <- emmeans(owls.brmsZINB2,~ArrivalTime|FoodTreatment,at=owls.grid,type=
                      'response') %>% as.data.frame()
head(owls.new)
```
```{r plot}
ggplot(owls.new,aes(x=ArrivalTime,y=prob))+
  geom_line(aes(color=FoodTreatment))+
  geom_ribbon(aes(ymin=lower.HPD,ymax=upper.HPD,fill=FoodTreatment),alpha=0.2)+
  theme_bw()
# based on our graph, it seems like the satiated group were fairly consistent with NCalls over time, whereas deprived group called substantially more than satiated groups at low Arrival Time

emmeans(owls.brmsZINB2,~ArrivalTime|FoodTreatment)


# all in all this was quite a complicated example
```

# References