---
title: "Bayesian GLMM example 4"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
mckeon = read_csv('../data/mckeon.csv', trim_ws=TRUE)
glimpse(mckeon)
mckeon <- mutate(mckeon, SYMBIONT=factor(SYMBIONT,levels=c('none','crabs','shrimp','both')),BLOCK=factor(BLOCK))
```


# Exploratory data analysis

Model formula:
$$
y_i \sim{} \mathcal{N}(n, p_i)\\
ln\left(\frac{p_i}{1-p_1}\right) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of symbionts on the probability of the colony experiencing predation.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual coral colonies.

# EXPLORE data with PLOT
```{r exploratory plot}
ggplot(mckeon,aes(x=SYMBIONT,y=PREDATION))+
  geom_point()+
  facet_wrap(~BLOCK)
```
## random intercept
# FIT model
```{r fit bayesian hierarchical (additive random intercept) model,cache=TRUE}
mckeon.form <- bf(PREDATION|trials(1)~SYMBIONT+(1|BLOCK),family=binomial(link='logit'))
mckeon.brms <- brm(mckeon.form,data=mckeon,iter=5000,chains=3,warmup=2000,thin=5,refresh=0)
#data is binary
```

# VALIDATE MODEL
```{r check chain characteristics}
mckeon.brms.var <- get_variables(mckeon.brms)
mcmc_plot(mckeon.brms,type='trace',pars=mckeon.brms.var[1:5])     # check trace plots
mcmc_plot(mckeon.brms,type='acf_bar',pars=mckeon.brms.var[1:5])   # check autocorrelation
mcmc_plot(mckeon.brms,type='neff_hist',pars=mckeon.brms.var[1:5]) # check effective samples
mcmc_plot(mckeon.brms,type='rhat_hist',pars=mckeon.brms.var[1:5]) # check chain convergence

# plots all look good so chains are well mixed and have converged on a stable posterior
```
```{r check overdispersion}
preds <- posterior_predict(mckeon.brms,nsamples=250,summary=FALSE)
mckeon.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = mckeon$PREDATION,
                             fittedPredictedResponse = apply(preds,2,median))
plot(mckeon.resids)

# slight issue of overdispersion: dispersion test p=0.032 (i.e. <0.05)
# residual plot of box plots: we want to see boxes that are symmetrical around the middle

# dispersion with binomial distribution - NOT easy to deal with => beta binomial (mixture) is able to model the variance rather than assume it, however because it does rely on the beta part it requires more data - with this dataset it's probably not big enough to fit a model using beta binomial

# if you have a large data set & a problem with overdispersion when fitting a binomial, you should try a beta binomial

# the other reason it could be overdispersed is that the model is too simple - so perhaps try a random intercept random slope model
```
## random intercept random slope
# FIT model
```{r fit random intercept, random slope model, cache=TRUE}
mckeon.form1 <- bf(PREDATION|trials(1)~SYMBIONT+(SYMBIONT|BLOCK),family=binomial(link='logit'))
mckeon.brms1 <- brm(mckeon.form1,data=mckeon,iter=5000,chains=3,warmup=2000,thin=10,refresh=0,control=list(adapt_delta=0.99))
# whenever you include a random slope for a categorical it will take a lot longer to run

# 6 divergent transitions but that's ok for the number of samples it did
```
# VALIDATE model
```{r check chain characteristics}
mckeon.brms.var1 <- get_variables(mckeon.brms1)
mcmc_plot(mckeon.brms1,type='trace',pars=mckeon.brms.var1[1:14])     # check trace plots
mcmc_plot(mckeon.brms1,type='acf_bar',pars=mckeon.brms.var1[1:14])   # check autocorrelation
mcmc_plot(mckeon.brms1,type='neff_hist',pars=mckeon.brms.var1[1:14]) # check effective samples
mcmc_plot(mckeon.brms1,type='rhat_hist',pars=mckeon.brms.var1[1:14]) # check chain convergence

# looks good now
```

```{r check overdispersion}
preds <- posterior_predict(mckeon.brms1,nsamples=250,summary=FALSE)
mckeon.resids <- createDHARMa(simulatedResponse = t(preds),
                             observedResponse = mckeon$PREDATION,
                             fittedPredictedResponse = apply(preds,2,median))
plot(mckeon.resids)

# yep overdispersion issue has been fixed now! wooo
```


# INVESTIGATE model / TEST HYPOTHESES
```{r}
ggpredict(mckeon.brms1) %>% plot
```
```{r model summary}
tidyMCMC(mckeon.brms1$fit,conf.int=TRUE,conf.method='HPDinterval',ess=TRUE,rhat=TRUE)

# seems like chance of being predated is lower when both symbionts compared to none
```
```{r planned comparisons}
# how about some planned comparisons via emmeans?
emmeans(mckeon.brms1,pairwise~SYMBIONT,type='response') # because we're on a log odds scale we need to convert back to response scale 

# example interpretation of these results: the odds of being predated on are 21 times higher for none group than crab group
# remember : odds are (p/(1-p))
```

```{r planned comparisons expressed on response scale}
# alright so what about probabilities?
mckeon.em <- emmeans(mckeon.brms1,pairwise~SYMBIONT,type='link')$contrasts %>% 
  gather_emmeans_draws() %>%  # i.e. don't summarise yet to medians & HPDs
  mutate(PEff=exp(.value)) # now we do the back transform ourselves by adding a new variable PEff (the backtransformed results)
mckeon.em %>% group_by(contrast) %>% dplyr::select(contrast,PEff) %>% median_hdi
mckeon.em %>% group_by(contrast) %>% dplyr::select(contrast,PEff) %>%summarise(P=sum(PEff>1)/n())


mckeon.em1 <-emmeans(mckeon.brms1,~SYMBIONT,type='link') %>% gather_emmeans_draws()
head(mckeon.em1) # .value column is on log odds scale - estimate of each mean 
mckeon.em1 %>% mutate(P=plogis(.value)) %>% head # the P column now represents probabilities (P(predation))
# plogis converts log odds scale to a probability scale

mckeon.em1 %>% mutate(P=plogis(.value)) %>% median_hdci(P) # so this tells us the probabilities of being predated on for each symbiont


# hdi = highest density interval (tries to find quantiles - but if it's bimodal - will give two difference confidence ribbons)
# hdci = highest density central interval - gives the central one
```


# PREDICT & PLOT using model
```{r}
mckeon.em1.plot <- mckeon.em1 %>% mutate(P=plogis(.value)) %>% median_hdci(P)
ggplot(mckeon.em1.plot,aes(x=SYMBIONT,y=P))+
  geom_pointrange(aes(ymin=.lower,ymax=.upper),color='black')+
  theme_bw()
```
# References
