---
title: "Data wrangling"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Install and load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse) #for data wrangling
```

Link to the data transformation cheatsheet

https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf

Important data manipulation functions:

| Task                      | Function                    | Package   |
|---------------------------|-----------------------------|-----------|
| Sorting                   | `arrange()`                 | **dplyr** |
| Re-ordering factor levels | `factor()`                  | **base**  |
| Re-labelling              | `recode()`                  | **dplyr** |
| Re-naming columns         | `rename()`                  | **dplyr** |
| Filtering/subsetting      | `select()`                  | **dplyr** |
| Transformations           | `mutate()`                  | **dplyr** |
| Adding columns            | `mutate()`                  | **dplyr** |
| Re-shaping data           | `gather()`/`spread()`       | **tidyr** |
| Aggregating               | `summarize()`, `group_by()` | **dplyr** |
| Merging/joining           | `*_join()`                  | **dplyr** |
|                           |                             |           |

# Piping

Use %>% to pipe the output of one function to another function & so on & so forth
You can use pipe with anything so long as the first argument of the function you're piping to calls for a data frame


# Data files

```{r getData, results='markdown', eval=TRUE}
load(file='../data/manipulationDatasets.RData') # .. means go back a directory and then go into data; RData takes up less space than a csv for example
data.1 %>% head %>% summary # taking the table data.1 and applying the head function to it to show only the start of the table (first 6 rows) 
str(data.1) # str is a string representation of the object - tells you what sort of data is in each variable
glimpse(data.1) # glimpse tells you the class of each variable
```

Sorting data
Most useful in presenting data, less important with analysis
================
```{r sorting}
arrange(data.1, LAT)      # arrange takes the data.frame, then the variable you want to sort by -                                this example sorts data.1 by LAT (latitude)

# now we can do the same thing using the pipe!
data.1 %>%                # take the datafram 'data.1'
  arrange(LAT)            # use pipe to sort by Latitude

# to sort in descending order....
data.1 %>%
  arrange(-LAT)

# sort on two variables...(use , or &)
data.1 %>%                 # take the dataframe 'data.1'
  arrange(Cond, Temp)      # sort by Cond, then sort by Temp (you can also use & i.e. arrange(Cond & Temp))

# you could sort by whatever you want e.g. the sum of two variables:
data.1 %>%
  arrange(LAT+LONG)

# sort by Between and then Cond
data.1 %>%
  arrange(Between,Cond)

# sort by Cond & then the ratio of Temp to LAT
data.1 %>%
  arrange(Cond,Temp/LAT)
```


Manipulating factors
=======================
```{r manipulating factors}
# R sorts factors in alphabetical order - you can re-level & re-label though - this is useful when graphing

data.3 <- data.1      # create a copy of data.1 & call it something else
#data.3 <- levels(data.3$Cond)

data.3$Cond <- factor(data.3$Cond, levels=c('L','M','H'), labels=c("Low", "Medium","High")) # provide labels for your factors 
```


Subset columns
SELECTING COLUMNS
=================

## Regular expressions (regex)

https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf

```{r selecting_columns}
select(data.1, Between, Plot, Cond, Time, Temp) # select only the variables you want to view from the whole data frame

# sometimes select can have trouble because it refers to functions from different packages so good practise to use it this way:

dplyr::select(data.1, Between, Plot, Cond, Time, Temp) # i.e. tell R what package the function 'select' is coming from

# type in the function name i.e. 'select' into the console and R will tell you the namespace i.e. the package the function has come from

# can select everything BUT certain variables BUT you can't combine the two i.e. you can't do select(data.1, BETWEEN, -LAT)
data.1 %>% select(-LAT,-LONG) # select everything except LAT & LONG
```

```{r helper_functions}
# Helper functions - string matching is NOT case sensitive

# contains()
# ends_with()
# starts_with()
# matches()

select(data.1,contains('L')) # select the variables that contain an L
select(data.1, starts_with('L')) # select the variables that START with L
select(data.1, ends_with('L')) # select the variables that END with L
# regX searching - regular expression searching
#select(data.1, match('^T[a-z]m.'))  # 4 letters long, ^ must start with T, [] any of those letters, must have an m, . is any letter' 

select(data.1,Between:Temp) # select every column from Between to Temp (only useful if you know the order)

# R has lots of data sets already available
head(nasa) # e.g. the nasa dataset 
nasa %>% select(lat, long, starts_with("cloud")) %>% head()

# select allows you to change columns because they appear in the order you type them
select(data.1, Time, everything()) # put Time in the front, and then leave everything else where it is

tikus[1:10,c(1:3,76:77)] # just another dataset in R
# select rep, time & only the species that DON'T contain pora
tikus %>%
  select(rep,time,everything(),-contains('pora'))

dplyr::select(tikus,`Pocillopora damicornis`) # if variable name has a space in it, refer to is using back ticks ` (the key on top left of keyboard with the squiggle character)

# what if you just want the vector that corresponds to the variable you want (rather than the data frame) - use 'pull'
data.1 %>% 
  pull(Temp) # spits out a vector with just the variable Temp

# rename allows you to change the names of variables 
data.1 %>% 
  rename(Condition=Cond, Temperature=Temp)
```


Filtering
SELECTING ROWS
=============
```{r filtering rows}
data.1 %>% 
  filter(Cond=="H") # only show me the rows where Cond equals H

data.1 %>% 
  filter(Cond %in% c('H','M')) %>%     # %in% operator means OR i.e. is Cond in the set of c('H','M') so is Cond H or M
  arrange(Cond)                        # now sort by Cond

data.1 %>% 
  filter(Cond=='H'|Cond=='M')          # this does the same thing, but can get a bit clunky & long-winded if you have lots of conditions to filter by

# filtering using different logical statements

data.1 %>% 
  filter(Cond=='H'& Temp < 25) # filter for Cond equals H AND temp is less than 25

data.1 %>% 
  filter(Cond=='H'|Temp < 25) # filter for Cond equals H OR temp is less than 25

# let's  keep only the rows with Temp less than 20 and LAT greater than 20 or LONG less than 145

data.1 %>% 
  filter((Temp < 20) & (LAT > 20 | LONG < 145))

# can also filter based on position
slice(data.1,1:4) # just keep first 4 rows
slice(data.1,c(1:4,7)) # just keep first 4 rows & the 7th row 
# can be useful to pipe together arrange then slice

sample_n(data.1,10,replace=TRUE) # will give you 10 rows from your data frame at random

# examine the levels of the Cond factor
data.3 <- filter(data.1,Plot=='P1')
data.processed <- data.1 %>% 
  filter(Plot=="P1") %>% 
  droplevels # this makes sure the levels that you don't want i.e. P2, P3, P4 are dropped from the dataframe - should be routine that you droplevels when filtering

```


Adding columns - mutate
===========================
```{r adding columns}

# create a new variable from existing ones & add into dataframe

data.1 %>% 
  mutate(LL=LAT+LONG)   # add a new column to the data.1 dataframe that is the sum of LAT & LONG

data.1 %>% 
  mutate(logTemp=log(Temp))

data.1 %>% 
  mutate(MeanTemp=mean(Temp),cTemp=Temp-MeanTemp) # centre the data around the mean (i.e. standardise it - probably need to do this if assuming normal distribution)

data.1 %>%  
  mutate(cTemp=Temp-mean(Temp))  # this code odes the sample thing but won't keep mean(Temp) as an extra column

data.1 %>%  
  mutate(leadTemp=lead(Temp), lagTemp=lag(Temp)) # useful for temporal data

```


Summarising (aggregating) data
When you summarise you end up with 1 row
=================================
```{r summarise}
data.1 %>%  mutate(mean(Temp)) %>% nrow() # 12 rows

data.1 %>% summarise(mean(Temp)) # summarize also works - s or z it doesn't matter # 1 row

data.1 %>% summarise(Mean=mean(Temp),Variance=var(Temp)) # summarise by mean & variance - can summarise by anything you want

#across versions of summarise:

data.1 %>% 
  summarise(across(c(Temp,LAT),list(Mean=mean,Var=var))) # calculate the mean and variance across the variables Temp and LAT 

data.1 %>% 
  summarise(across(where(is.numeric),list(Mean=mean, Var=var))) # calculate the mean & variance across all numeric variables

data.1 %>% 
  summarise(across(where(is.numeric),mean),across(where(is.factor),length)) # calculate the mean for all the numeric variables & the length of all the factor variables

```

Piping
Take the output of one function and pipe it into the next function (instead of doing functions inside functions)
=======
```{r}
## %>%                     # 'shift control m' is the shortcut for the pipe function
```


Grouping (=aggregating)
Takes categorical variables & splits data according to those variables
=========================
```{r grouping}

data.1 %>%                     # 'shift control m' is the shortcut for the pipe function
  group_by(Between,Plot) %>%   # group the data by the variables Between & Plot 
  summarise(Mean=mean(Temp))   # then calculate the Mean temp for each of those groups

data.1 %>%                     
  group_by(Plot) %>% arrange(Temp)  
```


Reshaping data
=================

Wide format - different observations are in different rows - the format data is often collected & stored in
Long format - every observation is in its own row (this is assumed) - the format data is analysed in

going from wide to long format - used to be called 'melting' - now called 'gathering'
## Gathering
```{r gathering}
### don't worry about this section too much.....
#gather(Time,Count, Time.0:Time.2)
#OR
#gather(Time, Count, -Between, -Plot)

# gather(what is repeated)
```

## Spread
```{r spreading}
#the reverse of gather is a 'spread' 
#select(-Resp2) %>% spread(Within,Resp1)
```

Combining data
=================
```{r merging}
# merging data frames
head(data.bio) # missing S3
head(data.chem) # missing S7
inner_join(data.bio, data.chem) # innerjoin only keep rows that are matched by both - so since S3 & S7 don't have matches, they do not appear in the merged data frame

full_join(data.bio,data.chem) # keeps everything even if things don't match

left_join(data.bio,data.chem) # keeps things that match with the stuff on the left - so there's no S3 here
right_join(data.bio, data.chem) # keeps things that match with the stuff on the right - so there's no S7 here

left_join(data.bio,data.geo,by=c("Plot")) # the 'by' bit tells which variables to match on
```


Applied examples
===================
