---
title: "GLMM example 4"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations

Load the necessary libraries

```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(broom.mixed)
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(glmmTMB)
```

# Scenario

Someone did something for some reason.....

# Read in the data

```{r readData, results='markdown', eval=TRUE}
mckeon = read_csv('../data/mckeon.csv', trim_ws=TRUE)
glimpse(mckeon)

# data is about predation on corals
# treatment (thing they want to look at) is whether having a symbiont (e.g. crab or shrimp) can actually reduce predation
# so in a tank they either had a predator and, a crab, a shrimp, a crab & shrimp, or nothing
# wanted to record whether colony experienced any predation or not: 1 = predation, 0 = no predation
# measure each of those treatments on each coral colony

# BLOCKS are the tanks
# TREATMENT is the symbiont i.e. crab, shrimp, both, nothing
# ok so since each of the treatments are applied to each block so this is a randomised complete/block


# RESPONSE is predation or not 1,0

# first convert BLOCK & SYMBIONT to factors
mckeon <- mutate(mckeon,SYMBIONT=factor(SYMBIONT),BLOCK=factor(BLOCK))

# now, 'both' is the first group here, but more sensible to make 'none' the first group, so need to change order of the levels:
mckeon <- mckeon %>% mutate(mckeon,SYMBIONT=factor(SYMBIONT,levels=c('none','crabs','shrimp','both')),BLOCK=factor(BLOCK))
  
```


# Exploratory data analysis

Model formula:
$$
y_i \sim{} \mathcal{N}(n, p_i)\\
ln\left(\frac{p_i}{1-p_1}\right) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed and random effects parameters respectively 
and $\bf{X}$ is the model matrix representing the overall intercept and effects of symbionts on the probability of the colony experiencing predation.
$\bf{Z}$ represents a cell means model matrix for the random intercepts associated with individual coral colonies.

```{r intial plot}
# let's plot this separately for each block
# treatment is categorical
# response variable is binary 0 or 1

mckeon %>% ggplot(aes(x=SYMBIONT,y=PREDATION))+
  geom_point(position=position_jitter(width=0.2, height=0))+
  facet_wrap(~BLOCK)

# you can see from these graphs, the likelihood of being predated is pretty high if no symbiont
# we also might expect that the corals respond differently - so worth looking at random intercept & slopes

# we know we're going to use a Binomial distribution because we've only got 0s & 1s
```


# Fit the model
```{r fit some models, see which one works best}
# fit random intercept model
mckeon_glmmTMB <- glmmTMB(PREDATION~SYMBIONT+(1|BLOCK),data=mckeon,family=binomial(link='logit'),REML=FALSE)

# fit random intercept slope model
mckeon_glmmTMB1 <- glmmTMB(PREDATION~SYMBIONT+(SYMBIONT|BLOCK),data=mckeon,family=binomial(link='logit'),REML=FALSE) # model didn't converge so can't use this

# only 1 mixed effect so don't need to worry about interactions

# ok so now we know we're going to go ahead with the first model, random intercept only, so just update it has REML=TRUE
mckeon_glmmTMB <- update(mckeon_glmmTMB,REML=TRUE) # this is our model we will use
```

# Model validation
```{r validate model}
# look at diagnostic plots
plot_model(mckeon_glmmTMB,type='diag')

# Binomial residuals are really hard to interpret, so the plot_model  only gives 1 plot (i.e. doesn't give residuals plot) - jsut gives QQ normal plot 

# so this QQ normal plot doesn't look too bad
```
```{r}
# BUT we can use DHARMa to deal with tricky residuals!!!
install.packages('DHARMa')
library(DHARMa)
simulatedResiduals <- simulateResiduals(mckeon_glmmTMB,plot=TRUE)

# looks pretty good - simulated residuals given as box plot because categorical 
```

# Model investigation / hypothesis testing
```{r}
summary(mckeon_glmmTMB)

# NB remember the link function we used - means these results are on a log odds scale!

# probability of being predated in the no symbiont groups is 83 times higher than the probability of not being predated on (because exp(4.42)=83.096)

# the odds of being predated on for crabs are only 4% of 83 because (exp(-3.32)=0.036) - so odds of being predated have declined by 96% - so 3 times as likely to be predated, rather than 83 times as likely

# for the group that has both symbionts, exp(4.42-5.152)=0.48 are the odds of being predated on - less likely to be predated on

# both symbionts exp(-5.15)=0.0058 

tidy(mckeon_glmmTMB,effect='fixed',conf.int=TRUE,exponentiate=TRUE)

# what about if we wanted to know for example...the difference b/w having a crab and having a crab & a shrimp? i.e. different comparisons - well we do a Tukey's test, OR we make our own comparisons
```
```{r contrast matrix to make other comparisons}
# so what about defining our own comparisons?? Let's say we're interested in this:

          # Crab vs. Shrimp             1 vs. 2 symbionts                   None vs. symbionts
# None           0                               0                               1
# Crab           1                              -1/2                            -1/3
# Shrimp        -1                              -1/2                            -1/3
# Both           0                               1                              -1/3

(cmat=cbind(crab_vs_shrimp=c(0,1,-1,0),one_vs_both=c(0,-1/2,-1/2,1),symbiont=c(1,-1/3,-1/3,-1/3)))

# important that the positives only sum to 1 and the negatives only sum to 1 otherwise you end up with effect size bigger than they really are

# ok now we have to make sure the contrast matrix meets the conditions of indpendence & the p-1 rule (so we had 4 groups, so can only do 3 comparisons)
crossprod(cmat) # the off diagonals are all zero so yep all the comparisons are independent, and there's only 3 comparisons so we're good to go!
```

```{r comparisons}
emmeans(mckeon_glmmTMB,~SYMBIONT,contr=list(SYMBIONT=cmat),type='response') # so here we're making our own contrasts rather than doing a Tukey's test, and we're asking for the on the response scale

# let's look at the contrasts - NB we're comparing odds:
# crab vs. shrimp - odds of being predated by crabs are nearly 90% higher than they are for shrimp (this is coming from the 1.89
# odds of being predated on if 1 symbiont is present 0.22, so likelihood of being predated on with 1 compared to having both is 4.5 times more likely 1/0.22=4.5
# if symbiont present vs. none odds are 62 - this is significant
```
# Strength of relationship
```{r r squared}
r.squaredGLMM(mckeon_glmmTMB) # the delta version is just the updated method of calculating r squared values
# so explaining about 14.5% of the variability when we don't account for the treatment (i.e. the different tanks) - when we put the blocking in, we can explain around 84% of the variability
```

```{r summary plot}
emmeans(mckeon_glmmTMB,~SYMBIONT,type='response') %>% 
  as.data.frame() %>% 
  ggplot(aes(x=SYMBIONT,y=prob))+
  geom_pointrange(aes(ymin=lower.CL,ymax=upper.CL))
```

# Predictions

# Summary figures

# References
