---
title: "Introduction to grammar of graphics"
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    css: ../resources/style.css
  pdf_document:
    df_print: default
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  word_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 4
    highlight: tango
    toc: yes
    toc_depth: 2
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r setup, include=FALSE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparations
Murray's tutorial notes pres 5.2 in docs folder
Load the necessary libraries
```{r libraries, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
# some useful packages for graphing
library(tidyverse) # for data wrangling
library(grid)
library(scales) # colour, size, axes etc. 
library(gridExtra)
```

Graphics infrastructure - essential elements of a graph
========================

- **layers** of data driven objects - at least 1 layer of data 
- **coordinate** system 
- **scales** - scale, legend, caption
- **faceting** - multiple graphs - should look alike
- **themes** - overall scheme of the graph - colour, size, font etc. etc. - aesthetics


Motivating data
-----------------
```{r, results='markup'}
head(BOD)
summary(BOD)
```

making graphs the old way
-------------------------
```{r}
# this is the long-hand version of making plots with ggplot but we never do it like this!!!! too long & messy
p <- ggplot() + # single layer - points
  layer(data=BOD, # data.frame
    mapping=aes(y=demand,x=Time),
    stat="identity", # use original data - identity means multiple the data point by 1 i.e. don't do anything to the raw data - plot as is
    geom="point", # plot data as points - geom says what geometric layer you want the data to display as
    position="identity", # where to place the points on the graph
    params = list(na.rm = TRUE),
    show.legend = FALSE # should there be a legend associated with this layer? e.g. axes scale, key etc.
  )+ # layer of lines - let's put another layer on our graph - a line!
  layer( data=BOD, #data.frame
    mapping=aes(y=demand,x=Time), # specific coordinates
    stat="identity", # use original data
    geom="line", # plot data as a line
    position="identity",
    params = list(na.rm = TRUE),
    show.legend = FALSE
  ) +
  coord_cartesian() + #cartesian coordinates
  scale_x_continuous() + #continuous x axis
  scale_y_continuous()  #continuous y axis
```
making graphs the new way
-------------------------
```{r}
# short-hand version - much easier
ggplot(data=BOD, map=aes(y=demand,x=Time)) +  # aes means aesthetics
  geom_point()+
  geom_line()
# since the first argument in the ggplot function required is the data frame we can pipe our data to ggplot:
BOD %>% ggplot(map=aes(y=demand,x=Time))+
                 geom_point()+ # refer to layer by its geometric feature i.e. point
                 geom_line() # these additional layers inherit all the information from the ggplot function - but you can state separately
```

Layers
==========

- **layers** of data driven objects
    - **geom**etric objects to represent data
    - **stat**istical methods to summarize the data
    - **mapping** of aesthetics
    - **position** control

Primary geometric objects
===========================

`geom_bar`
------------
# Single continuous variable - HISTOGRAM
```{r}
head(diamonds) # example data
ggplot(diamonds) + geom_bar(aes(x = carat)) # when you do geom_bar and you give it a continuous variable you get a histogram - needs to bin the data which you can specify if you want but generally don't
```
# Single categorical variable - BAR CHART
```{r}
ggplot(diamonds) + geom_bar(aes(x = cut))
```
# Multiple categorical variables - STACKED BAR CHART IS DEFAULT
```{r}
ggplot(diamonds) + geom_bar(aes(x = cut, fill = clarity)) # multiple categorical variables are stacked by default
# if you changed coordinate system to polar coordinates you'd get a pie chart!

ggplot(diamonds) + geom_bar(aes(x = cut, fill = clarity),
  position='dodge') # conditional bar char - bars are contingent on clarity - no longer stacked
```

`geom_boxplot`
----------------
good way to quickly visualise continuous data i.e. outliers, skewness etc.

# univariate box plot
```{r}
ggplot(diamonds) + geom_boxplot(aes(x = "carat", y = carat))
ggplot(diamonds) + geom_boxplot(aes(y = carat)) # this does the same thing - don't need to specify the x variable
```
# conditional box plot
```{r}
ggplot(diamonds) + geom_boxplot(aes(x = cut, y = carat)) # multiple different categories
```

`geom_line`
-------------
```{r}
ggplot(CO2) + geom_line(aes(x = conc, y = uptake)) # this looks like nonsense - let's plot things separately according to a grouping variable e.g. 'plant'
```
# Grouping
```{r}
ggplot(CO2) + geom_line(aes(x = conc, y = uptake, group=Plant)) # this is much better - we have plotted each group separately
```
# Colour Scale
```{r}
ggplot(CO2) + geom_line(aes(x = conc, y = uptake, colour=Plant)) # plot separately and colour each plot with a different colour & this also applies a legend! - standard now to represent things by two different scales to satisfy visual cues i.e. colour & shape, colour & line type
```

`geom_point`
------------
# Shape & Colour scale based on categorial variables in the data
```{r}
ggplot(CO2) + geom_point(aes(x = conc, y = uptake, colour=Treatment, shape=Treatment))
ggplot(CO2) + geom_point(aes(x = conc, y = uptake, colour=Plant, shape=Type))
glimpse(CO2)
```

`geom_smooth`
------------
# Useful as a diagnostic tool when contemplating what model to fit the data to BUT not appropriate for statistical analysis!!!!
```{r}
ggplot(CO2) + geom_smooth(aes(x = conc, y = uptake), method='lm') # runs a statistical model to fit a smoother to your data in the background (in this case the method is lm i.e. linear model) & plotting line of best fit & confidence intervals

# now add data to the smoothed plot
ggplot(CO2) + geom_smooth(aes(x = conc, y = uptake), method='lm') +
  geom_point(aes(x=conc,y=uptake)) # we can see a linear model is not the best option for this dataset - so geom_smooth can be a good quick diagnostic tool when initially looking at your data to see what model might be best to approximate the data
```
# Smooth by grouping variable
```{r}
# what if we smoothed by a grouping variable - this might help visualise the patterns??
ggplot(CO2) + geom_smooth(aes(x = conc, y = uptake, group=Plant), method='lm') +
  geom_point(aes(x=conc,y=uptake))
```
# Smooth using a GAM instead of lm (i.e. different smoothing method)
```{r}
ggplot(CO2) + geom_smooth(aes(x = conc, y = uptake, group=Plant)) +
  geom_point(aes(x=conc,y=uptake)) # by not specifying the method, ggplot defaults to GAM (generalised linear model) as the model to fit
```

`geom_polygon`
----------------
# Making maps - R has some good packages for this - sort of derived from GIS

`geom_tile`
----------------

`geom_raster`
----------------

EXAMPLES - GEOM_POINT
----------------------
# Store instructions for making a graph in ggplot
```{r}
p <- ggplot(data=BOD)
p <- p + geom_point(aes(y=demand,x=Time))
p # this is doing print(p) - need to write print(p) if you're writing it in a function
```

Secondary geometric objects
============================

`geom_errorbar`
------------------

`geom_pointrange`
------------------

Coordinate systems
====================
# Default in R: cartesion coordinates
```{r}
ggplot(CO2)+geom_point(aes(x=conc,y=uptake))+
  coord_cartesian() 
```
# Flip x & y
```{r}
ggplot(CO2)+geom_point(aes(x=conc,y=uptake))+
 coord_flip()
```
# Polar coordinates - good for cyclical data
```{r}
ggplot(CO2)+geom_point(aes(x=conc,y=uptake))+
  coord_polar()
```
# Orthographic coordinates - can change projections & reference system for maps

Scales
==========

`scale_x_` and `scale_y_`
----------------------------
# Naming axes
```{r}
ggplot(CO2, aes(y=uptake,x=conc)) + geom_point()+
  scale_x_continuous(name="CO2 conc") # what about superscripts? mathematical notation? greek letters?? we need a better way to name axes!

# we need to use an expression to do this - NB cannot have spaces in an expression!:
ggplot(CO2, aes(y=uptake,x=conc)) + geom_point()+
    scale_x_continuous(name=expression(
      Ambient~CO[2]~concentration~(mg/l))) # ~ is a space, [] is subscript, ^ is superscript
# reference for how to write different characters - type demo(plotmath) into console and press enter to see all the different ways to get symbols
```
# Expanding axes
```{r}
ggplot(CO2, aes(y=uptake,x=conc)) + geom_point()+
  scale_x_continuous(name="CO2 conc", expand=c(0,200)) # c(multiplier, addition) c(0,200) means 0 don't mulitiply scale but add 200 units - i.e. how to scale axes
```
# Axis on a log scale
```{r}
ggplot(CO2, aes(y=uptake,x=conc)) + geom_point()+
  scale_x_log10(name="CO2 conc",
    breaks=as.vector(c(1,2,5,10) %o% 10^(-1:2))) # breaks specifies where to put the tick marks & labels on the x axis 
# 10^(-1:2) creates a vector of integers from 10^-1 to 10^2
# %o% operator means calculate the outer product so c(1,2,5,10) %o% 10^(-1:2)) gives you the outer product of c(1,2,5,10) & c(0.1,1,10,100) - we want to make the matrix result a vector so provide as.vector and that will give the vector required for where to put the tick marks

# if you change to log scale or square root scale you should identify the breaks / tick marks so it's more sensible
```
# Scale categorical data
```{r}
ggplot(CO2, aes(y=uptake,x=Treatment)) + geom_point()+
  scale_x_discrete(name="Treatment")
```
Other scales
---------------

- **size** of points (thickness of lines)
- **shape** of points
- **linetype** of lines
- **color** of lines or points
- **fill** of shapes

`scale_size`
----------------

`scale_shape`
----------------

`scale_linetype`
----------------

`scale_fill` and `scale_color`
-------------------------------

EXAMPLES - SCALES
---------------
```{r}
(p <- p + scale_x_sqrt(name="Time")) # changes scale of axes (doesn't change the data) # NB you should plot raw data on transformed scales, not transformed data on linear scale - this is important! easier to read the graph & interpret raw data
(p <- p + scale_x_log10(name="Time"))
```

Facets
==========
# matrices of plots
# facet_wrap
```{r}
ggplot(CO2,aes(x=conc,y=uptake, colour=Type))+
  geom_point()+   
  facet_wrap(~Plant)+    # facet on the plant - we want a different plont for each plant type
  geom_smooth()    # smooth each graph

# only time order matters is when each layer is drawn so smooth before points means the gray smooth is drawn before the points
```
# Change scale for each graph in facet
```{r}
ggplot(CO2)+geom_line(aes(x=conc,y=uptake, colour=Type))+ 
facet_wrap(~Plant, scales='free') # free scales means the scales appear on every single graph
ggplot(CO2)+geom_line(aes(x=conc,y=uptake, colour=Type))+ 
facet_wrap(~Plant, scales='free_y') # free scale only for y axis, free_x will do free axes only for x axis
```
# facet_grid
```{r}
ggplot(CO2)+geom_point(aes(x=conc,y=uptake, colour=Type))+ 
facet_grid(Type~Treatment) # MUST have common x & y axes - can't differ axes like you can in facet_wrap
```

Themes
==========
Default theme in R is theme_grey()
type in theme_classic() into console to see all of the specified features - then you can see how to change certain bits

`theme_classic`
----------------
```{r}
ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_linedraw() # just changes a bunch of stuff to make the graph look different which you can type out and code manually but these are just pre-packaged themes - can start with a theme and then make changes on top of that
```
`theme_bw`
----------------
```{r}
ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_bw()
```
`theme_grey`
----------------
```{r}
ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_grey()
```
`theme_minimal`
----------------
```{r}
ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_minimal()
```
`theme_linedraw`
----------------
```{r}
ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_linedraw()
```
# changing specific parts of a theme
```{r}
ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_classic() + theme(axis.title.x=element_text(margin=margin(t=2,unit='pt')))
```
Practice
===========

EXAMPLES - GEOM_POINT WITH OPTIONAL MAPPING
--------------------------------------------
# Colour
```{r}
ggplot(CO2) + 
  geom_point(aes(x=conc,y=uptake), colour="red")      # colour can be coded with words

ggplot(CO2) + 
  geom_point(aes(x=conc,y=uptake), color="#ff0000")   # colour can be coded with a hash code - refer to R colour chart for hash codes

# but there's no scale saying what the red dots mean!
```
# Aesthetics & Scale
```{r}
ggplot(CO2) + 
  geom_point(aes(x=conc,y=uptake, color="blue")) # anything INSIDE the aesthetic has a scale - if it's OUTSIDE the aesthetic it DOESN'T have a scale - if it's outside it's treated literally, if it's inside the brackets it's not treated literally, it says colour it according to what groups I have
```
# Graphing different statistics - input raw data, plot summary data
```{r}
ggplot(CO2)+geom_point(aes(x=conc,y=uptake),
 stat="summary",fun.y=mean) # adds mean to the graph - passing raw data and plotting aggregated data - this isn't the best way to do things - you should aggregate your data first then pass directly to the graph
```
Exporting graphs
================
# how to export a pretty graph from R??????
```{r}
# first store graph
g1 <- ggplot(CO2, aes(y = uptake, x = conc)) + geom_smooth() +
    geom_point() + theme_linedraw()
ggsave(filename='/Users/sfulton/OneDrive - James Cook University/JCU/R/AIMSJCU ML R COURSE/ML R COURSE/testgraph.png', g1, width=5, height=5, units='cm',dpi=300) # default unit for size are in inches so can specify units if need be # can set resolution also

# good to have a folder called output and then others inside that called tables, figures etc. 
```
